Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
Minor fixes
Add an example of decomposed KS1 proof

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a ``syntax-free'' notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula$ is depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%
\begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
  \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
  \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
\end{tikzpicture}\vspace{0ex}\end{center}
%
Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

@@

@@

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1,\dots,t_n$ or $pt_1,\dots,t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

The obvious next step in our research is to investigate proof
composition and normalisation of first-order combinatorial
proofs. Even in the propositional setting, the normalization of
combinatorial proofs is underdevelopped. There exist two different
procedures for cut elimination~\cite{hughes:invar,str:FSCD17}, but
both have their insufficiencies, and there is no general theory.

>>>>

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a ``syntax-free'' notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula$ is depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%
\begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
  \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
  \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
\end{tikzpicture}\vspace{0ex}\end{center}
%

\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqn{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}}{
    \vlin{\conr}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p}}{
      \vliin{\vlan}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p, p}}{
        \vlin{\vlor}{}{\sqn{\cneg{p} \vlor q, p}}{
          \vlin{\weakr}{}{\sqn{\cneg{p}, q, p}}{
            \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
              \vlhy{}}}}}{
        \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
          \vlhy{}}}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cD}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}{
    \vlin{\fequ}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor (p \vlor p)}{
      \vlin{\sw}{}{(\cneg{p} \vlan (\cneg{p} \vlor q)) \vlor p) \vlor p}{
        \vlin{\fequ}{}{(\cneg{p} \vlan ((\cneg{p} \vlor q) \vlor p)) \vlor p)}{
          \vlin{\sw}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \cneg{p}) \vlor p}{
            \vlin{\aiD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan (\cneg{p} \vlor p)}{
              \vlin{\tttD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \ttt}{
                \vlin{\fequ}{}{p \vlor (\cneg{p} \vlor q)}{
                  \vlin{\wrD}{}{(p \vlor \cneg{p}) \vlor q}{
                    \vlin{\aiD}{}{p \vlor \cneg{p}}{
                      \vlhy{\ttt}}}}}}}}}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py))}}{
    \vlin{\exists}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py)),\exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
      \vlin{\vlor}{}{\sqn{\cneg{p}w \vlor (\forall y.py), \exists
x.(\cneg{p}x \vlor (\forall y.p(y)))}}{
        \vlin{\forall}{}{\sqn{\cneg{p}w, \forall y.py, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
          \vlin{\exists}{}{\sqn{\cneg{p}w, pz, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
            \vlin{\vlor}{}{\sqn{\cneg{p}w, pz, \cneg{p}z \vlor (\forall y.py))}}{
              \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z, (\forall y.py))}}{
                \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z}}{
                \vlin{\axr}{}{\sqn{pz, \cneg{p}z}}{
                  \vlhy{}}}}}}}}}}}
\end{equation*}
\juihsuan{LK1 and KS1 proofs of Peirce's law and LK1 proof of drinker formula}

Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

@@

\begin{equation*}
\vlderivation{
  \vlin{\acD}{}{(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))}{
    \vlin{\mexD}{}{(\exists x. (\cneg{p}x \vlor \cneg{p}x) \vlor (\forall
y.(py \vlan pfy))}{
      \vlin{\fequ}{}{((\exists x. \cneg{p}x) \vlor (\exists x.\cneg{p}x))
\vlor (\forall y.(py \vlan pfy)}{
        \vlin{\exists}{}{\forall y.(((\exists x. \cneg{p}x) \vlor (\exists
x.\cneg{p}x)) \vlor (py \vlan pfy))}{
          \vlin{\exists}{}{\forall y.((\cneg{p}y \vlor (\exists x.\cneg{p}x))
\vlor (py \vlan pfy))}{
            \vlin{\fequ}{}{\forall y.((\cneg{p}y \vlor \cneg{p}fy) \vlor
(py \vlan pfy))}{
              \vlin{\sw}{}{\forall y.(\cneg{p}y \vlor ((py \vlan pfy)
\vlor \cneg{p}fy))}{
                \vlin{\fequ}{}{\forall y.(\cneg{p}y \vlor (py \vlan (pfy
\vlor \cneg{p}fy)))}{
                  \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan (pfy
\vlor \cneg{p}fy))}{
                    \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan
\ttt)}{
                      \vlin{\ttt}{}{\forall y.(\ttt \vlan \ttt)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
\end{equation*}
\juihsuan{the decomposed KS1 proof of $(\exists x. \cneg{p}(x)) \vlor (\forall
y.(p(y) \vlan p(f(y))))$}

@@

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

The obvious next step in our research is to investigate proof
composition and normalisation of first-order combinatorial
proofs. Even in the propositional setting, the normalization of
combinatorial proofs is underdevelopped. There exist two different
procedures for cut elimination~\cite{hughes:invar,str:fscd17}, but
both have their insufficiencies, and there is no general theory.
